---
title: "Introducing ggplot"
output: html_notebook
date: 29/10/2021
author: Martin Shepperd
---

# Introducing ggplot

This is a short guide to help provide a feel for using ggplot2 and tobecome aware of the almost limitless possibilities.  

```{r}
# This assumes {ggplot2} and {ploty} have already been installed
library(ggplot2)
library(plotly)     # This package provides mouseover functionality to ggplots

```

To start with, we need some data so let's use the built in data set mtcars.

```{r}
# Load and view the data
data("mtcars")
View(mtcars)

# Eyeball the structure
str(mtcars)
```


```{r}
# First identify the data
p1 <- ggplot(mtcars, aes(x=hp, y=mpg))
p1   # NB This plot isn't at all interesting as we haven't plotted anything (yet)!

# Add points for scatterplot
p1 <- p1 +
      geom_point(col="steelblue", shape=3)   # Shape modifies the data point
p1

# Add a linear regression line using the lm() function
p1 <- p1 +
      geom_smooth(method = "lm")  
p1

# Add some meaningful titles to improve understandability
p1 <- p1 +
      ggtitle("Horsepower vs Fuel Consumption", subtitle="From the mtcars dataset") + 
      xlab("Horse power") +
      ylab("Miles per gallon")
# View the p1 plot object
p1
```

### Using {plotly}

This package contains the ggploty() which adds a mouseover feature to ggplots.  Sometimes it's helpful to know the exact values of a data point.  Anyhow, I think it's quite cute!

```{r}
# Apply the function to the p1 ggplot object which has just been created.
ggplotly(p1)
```

We can continue to play with the chart, by changing the regression line to a loess smoother hence we no longer assume a linear relationshp between hp and mpg.  Note that for high hp values the pattern becomes weak and the confidence intervals widen accordingly.

```{r}
# Or to add extra information we can additionally display row names (ie the car type)
p2 <- ggplot(mtcars, aes(x=hp, y=mpg, text = paste("Car:", rownames(mtcars)))) +
      geom_point(col="steelblue", shape=3) +
      geom_smooth(method = "loess", se = T) +
      ggtitle("Horsepower vs Fuel Consumption", subtitle="From the mtcars dataset") + 
      xlab("Horse power") +
      ylab("Miles per gallon")
ggplotly(p2)
```


### Adding extra dimensions

Let's enrich our visual analysis with a third variable `am` which represents transmission type.  We might hypothesise that manual cars will be more fuel efficient which is borne out by the plot below.  NB I strip out some of the previous features such as changing the colour of the data points, so we can follow the next tranche of additions.  

We note from the `str()` function that all variables are numeric even though am is actually a grouping factor and the 0 and 1 values are entirely arbitrary.  To use it meaningfully we need to turn it into a **factor**.  Also we need to check the meta-data to understand how the levels are coded (as 0 and 1).  

```{r}
# Add a 3rd variable (weight) to analysis, ie 
# do automatic cars have worse fuel consumption?
# First turn am into a factor
mtcars$am <- as.factor(mtcars$am)

# Use colour of data points to different transmission types
p3 <- ggplot(mtcars, aes(x=hp, y=mpg)) +
      geom_point(shape=3, aes(color=am)) +
      geom_smooth(method = "loess") +       # Change from a linear regression to a loess smoother
      ggtitle("Horsepower vs Fuel Consumption", subtitle="From the mtcars dataset") + 
      xlab("Horse power") +
      ylab("Miles per gallon") +
      scale_color_discrete(name="Transmission",
                       labels=c("Automatic","Manual"))
p3
```
