---
title: "ConvertCumulativeToNew"
output: html_notebook
author: Martin Shepperd
date: 13/10/2020
---

This short R Notebook seeks to explain how the cumulative counts for covid-19 mortality and infection can easily be converted into new counts by processing the vectors.

I will show the functioning for an extremely simplified case of 5 days.

```{r}
# Make a vector of cumulative infections
cumInf <- c(0,1,1,4,6)

# Turn it into a dataframe for easy viewing
df1 <- as.data.frame(cumInf)
```

Now we want to subtract the previous day's infection from today's to give the count of new infections.  You could do this with a loop and sequence from start to end of the vector, but *remember* R is a functional language and has powerful vector processing capabilities.

```{r}
# The head function head(x,n) returns n elements from x
# so -1 returns the entire vector less one
head(cumInf,-1)

# Likewise with tail, except starting from the bottom
tail(cumInf,-1)
```

So we can put all this together in one neat statement to turn cumulative counts into new counts.

```{r}
newInf <- tail(cumInf,-1) - head(cumInf,-1)
newInf
```

But we lack one thing, which is the first element.  Here we only have n-1 values, the first day's count is missing because effectively we need a day[0]=0.  So we need to add day[1] where the new count *is* the cumulative count, since there is no previous count.

```{r}
newInf <- c(cumInf[1],newInf)
```

And finally to add the new infections to the existing data frame df1

```{r}
# Add a new column to df1 using the cbind() function
df1 <- cbind(df1, newInf)
```

